\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-02-01 latex/article_vol2.tex
Submitted version

\pkg{dataMaid} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{dataMaid} is fully
customizable and \code{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{dataMaid} output file
can be anything --- depending on the arguments given to \code{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously.
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \code{summaryFunction}s,
\code{visualFunction}s and \code{checkFunction}s are made.

This section consists of two parts. First, we describe how customized \code{summaryFunction}s,
\code{visualFunction}s and \code{checkFunctions} can be made, including an introduction to a
few convenient tools available in \pkg{dataMaid} for aiding SVC-function construction.
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \code{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} - a variable vector \newline \code{...}  &  \code{v} - a variable vector \newline \code{vnam} - the variable name (as character string) \newline \code{doEval} - a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...}  \\
Input (optional) &  \code{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \code{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} - a label for the summary value (as character string); \newline \code{\$result} - the result of the summary (as character string) & A character string with \proglang{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} - a logical identifying whether an issue was found; \newline \code{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \code{rmarkdown} \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \proglang{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \proglang{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary-, visual or check function, one needs
to create a new function with a specific structure.  This can be done with
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \code{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{dataMaid}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{dataMaid}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

We will now build four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID()}] A new \code{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual()}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros()}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons()}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

\subsubsection{isID --- a new checkFunction without problem values}
First, let's define the \code{isID()} function. As this function is not
supposed to list problematic values, it falls within
the category of \code{checkFunction}s represented by
\code{myFullVarCheckFunction()} in the above. We do not particularly wish
to use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

\begin{Schunk}
\begin{Sinput}
R> isID <- function(v, nMax = NULL, ...) {
+    out <- list(problem = FALSE, message = "")
+    # Should work for all but logical and Date variables
+    if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
+      v <- as.character(v)
+      lengths <- c(nchar(v))
+      # Check that all lengths are the same and greater than 10 characters
+      if (all(lengths > 10) & length(unique(lengths)) == 1) {
+        out$problem <- TRUE
+        out$message <- "Warning: This variable seems to contain ID codes."
+      }
+    }
+    out
+  }
\end{Sinput}
\end{Schunk}

This is essentially all we need to do in order to include this
function as a precheck-function in \code{clean()}, so we will leave it as
is and move on to the next function, namely \code{mosaicVisual()}.

\subsubsection{mosiacVisual --- a new visualFunction}
\proglang{R} contains a function, \code{mosiacplot()}, which produces
mosaic plots. We intend to use this existing high-level plotting
function as part of the new visualization function.  We will define
the new function such that it gets the full \pkg{dataMaid}
functionality. This can be done using the following code which sets up
the call using the existing function \code{mosaicplot()}.

\begin{Schunk}
\begin{Sinput}
R> mosaicVisual <- function(v, vnam, doEval) {
+    thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
+    if (doEval) {                     # Return the graphics
+      return(eval(thisCall))
+    } else return(deparse(thisCall))  # Else return the code for the call
+  }
\end{Sinput}
\end{Schunk}

@@

\subsubsection{Economic memory use}
Another solution, which is
especially relevant to Windows users due to %the unfortunate
the memory handling on this operating system,
%combination of memory control in this operating system and
%\proglang{R},
is simply splitting the two steps performed
by \code{clean()}, namely producing the \pkg{rmarkdown} file and rendering it
afterwards. If the \code{rmarkdown} file is very long, as it will
typically be for very large datasets, keeping this file open in memory
wastes precious memory capacities. Therefore, we advice users to
instead split the two steps. This can be done in the following manner:

@@

This also deals with the fact that \pkg{dataMaid} can produce
\pkg{rmarkdown} files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the human time it takes to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. Or even for more reasonably sized
datasets, sometimes a brief overview of the most pressing issues can
be useful. This can be achieved by using the \code{onlyProblematic}
argument in \code{clean()}. By specifying \code{onlyProblematic = TRUE},
only variables that raise a flag in the checking steps will be
summarized and visualized. But perhaps we are not even interested in
obtaining general information about these variables, but only in
getting a quick overview of the problems they might have. This can be
done by also controlling the \code{mode} argument:

@@

\subsection[Including dataMaid documents in other files]{Including \pkg{dataMaid} documents in other files}
Sometimes, a \pkg{dataMaid} document might be a useful addition to a
more general overview document, including also for example pairwise
association plots, time series plots, or exploratory analysis
results. To this end, it is possible to produce a \pkg{dataMaid}
document that can readily be included in other \pkg{rmarkdown}
files. This is done by using the \code{standAlone} argument in \code{clean()},
which removes the preamble from the outputted \pkg{rmarkdown}
file. Please note, that it is still necessary to indicate which
\code{rmarkdown} type is being created; the pdf and html \pkg{rmarkdown}
styles are unfortunately not identical.

If it is important that the embedded \pkg{dataMaid} document can be
rendered to either of these two file types, we recommend setting
\code{twoCols = FALSE} and \code{output = html} in \code{clean()}, thereby
essentially removing almost all output type specific formatting code
from the generated \pkg{rmarkdown} file.

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \pkg{rmarkdown}
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \code{dataMaid\_toyData.Rmd} file can
then be included:

@@

and the more minimal, html-style \pkg{rmarkdown} file described above can be produced using

@@

Note that with the latter option, no special YAML preamble is needed in the \pkg{rmarkdown} document.
% \hl{don't end on code.}

>>>>

\pkg{dataMaid} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{dataMaid} is fully
customizable and \code{clean()} is mainly a tool for formatting the
results from various checking, summary,  and visualization
functions. Thus, the actual work underlying a \pkg{dataMaid} output file
can be anything --- depending on the arguments given to \code{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously.
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \code{summaryFunction}s,
\code{visualFunction}s and \code{checkFunction}s are made.

This section consists of two parts. First, we describe how customized
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunctions} can be made, including an introduction to a few
convenient tools available in \pkg{dataMaid} for aiding SVC-function
construction.  Table \ref{table.functionTypes} serves as an overview
of the internals of these three central function types. Secondly, we
turn to a worked example of how to use custom made functions in
practice; Four new SVC functions are defined and used, both
interactively and in \code{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} --- a variable vector.
                                \newline \code{...} --- additional
                                arguments passed to the function.  &  \code{v} --- a variable vector. \newline \code{vnam} --- the variable name (as character string). \newline \code{doEval} --- a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function. & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...} --- additional arguments passed to the function. \\
Input (optional) &  \code{maxDecimals} --- number of decimals printed in outputted numerical values.  & --- &  \code{maxDecimals}  --- number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} --- a label for the summary value (as character string); \newline \code{\$result} --- the result of the summary (as character string). & A character string with \proglang{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} --- a logical identifying whether an issue was found; \newline \code{\$message} --- a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use with \code{rmarkdown}. \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \proglang{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \proglang{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the summarize, visualize, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary, visual or check function, one needs
to create a new function with a specific structure.  This can be done with
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \code{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{dataMaid}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{dataMaid}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

As an example, we create four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID()}] A new \code{checkFunction} intended for use in the pre-check-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual()}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros()}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons()}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

\subsubsection[isID --- a new checkFunction without problem values]{\code{isID} --- a new \code{checkFunction} without problem values}
First, let's define the \code{isID()} function. As this function is
not supposed to list problematic values, it falls within the category
of \code{checkFunction}s represented by
\code{myFullVarCheckFunction()} above. We do not particularly wish to
use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

\begin{Schunk}
\begin{Sinput}
R> isID <- function(v, nMax = NULL, ...) {
+    out <- list(problem = FALSE, message = "")
+    if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
+      v <- as.character(v)
+      lengths <- c(nchar(v))
+      if (all(lengths > 10) & length(unique(lengths)) == 1) {
+        out$problem <- TRUE
+        out$message <- "Warning: This variable seems to contain ID codes."
+      }
+    }
+    out
+  }
\end{Sinput}
\end{Schunk}

This is essentially all we need to do in order to include this
function as a pre-check-function in \code{clean()}, so we will leave it as
is and move on to the next function, namely \code{mosaicVisual()}.

\subsubsection[mosaicVisual --- a new visualFunction]{\code{mosaicVisual} --- a new \code{visualFunction}}
\proglang{R} provides a function, \code{mosaicplot()}, which produces
mosaic plots. We intend to use this existing high-level plotting
function as part of the new visualization function.  We will define
the new function such that it gets the full \pkg{dataMaid}
functionality. This can be done using the following code which sets up
the call using the existing function \code{mosaicplot()}.

\begin{Schunk}
\begin{Sinput}
R> mosaicVisual <- function(v, vnam, doEval) {
+    thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
+    if (doEval) {
+      return(eval(thisCall))
+    } else return(deparse(thisCall))
+  }
\end{Sinput}
\end{Schunk}

@@

\subsubsection{Economic memory use}
Another solution, which is
especially relevant to Windows users due to %the unfortunate
the memory handling on this operating system,
%combination of memory control in this operating system and
%\proglang{R},
is simply splitting the two steps performed
by \code{clean()}, namely producing the \proglang{R} markdown file and rendering it
afterwards. If the \code{rmarkdown} file is very long, as it will
typically be for very large datasets, keeping this file open in memory
wastes precious memory capacities. Therefore, we advice users to
instead split the two steps as shown in the following.

@@

This also deals with the fact that \pkg{dataMaid} can produce
\proglang{R} markdown files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the time it takes a human to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. % Or even for more reasonably sized
% datasets, sometimes a brief overview of the most pressing issues can
% be useful.
This can be achieved by using the \code{onlyProblematic} argument in
\code{clean()}. By specifying \code{onlyProblematic = TRUE}, only
variables that raise a flag in the checking steps will be summarized
and visualized. But perhaps we are not even interested in obtaining
general information about these variables, but only in getting a quick
overview of the problems they might have. This is obtained by
controlling the \code{mode} argument:

@@

\subsection[Including dataMaid documents in other files]{Including \pkg{dataMaid} documents in other files}
Sometimes, a \pkg{dataMaid} document might be a useful addition to a
more general overview document, including additional information such
as pairwise association plots, time series plots, or exploratory
analysis results.  \pkg{dataMaid} can produce a document to be
included in other \proglang{R} markdown files by setting the
\code{standAlone} argument in \code{clean()} to remove the preamble
from the output \proglang{R} markdown file. Note that it is still
necessary to indicate which \proglang{R} markdown output type is
created; the pdf and html \proglang{R} markdown styles are
unfortunately not identical.

If it is important that the embedded \pkg{dataMaid} document can be
rendered to either of these two file types, we recommend setting
\code{twoCols = FALSE} and \code{output = "html"} in \code{clean()}, thereby
essentially removing almost all output type specific formatting code
from the generated \proglang{R} markdown file.

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \proglang{R} markdown
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \code{dataMaid\_toyData.Rmd} file can
then be included:

@@

and the more minimal, html-style \proglang{R} markdown file described above can be produced using

@@

Note that with the latter option, no special YAML preamble is needed in the \proglang{R} markdown document.
% \hl{don't end on code.}
